R Notebook

### 2022-04-25 article/abs.Rmd
draft res,fig3 bug

This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and
addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolorant or late successional taxa, and that habitat edges affect regenerating community composition.

@@

>>>>

This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge in a former timber plantation context, and
addresses the hypothesis that old plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and that habitat edges affect regenerating community composition.

@@

# SACNAS

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, medaited by shade tolerance and dispersal traits, along with management history. 
This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition. 
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature. 
As distance from primary forest edge increased, biomass and stem and wood density tended to increase, with ca 10% variation explained, while canopy light and stand height tended to stay the same. 
Stand tree richness also tended to increase, but diversity decreased steeply and non-linearly, and composition varied notably. 
Finally, tree taxa associated with both early and late successional stages increased, but biomass by dispersal mode did not tend to change. 
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

### 2022-04-25 article/intro.Rmd
draft res,fig3 bug

---
title: "Introduction"
output: html_notebook
---

1.  importance of reforestation/regeneration/regrowth

2.  patterns of reforestation

    1.  general

    2.  tropical

3.  things that affect reforestation

    1.  edge effects

    2.  plantation structure?

4.  [for NGO--start biomass census]

    1.  Does leaving the plantation structure help reforestation?

        1.  shade

        2.  edge effects (i.e. buffer)

    2.  How does the forest regenerate by distance to primary forest?

        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?

@@

-   Tropical forests store most terrestrial biomass ( @pan2013 ) and biodiversity (CITE, but are losing functions, especially in the neotropics ( @hubau2020 ), as most tropical forests become secondary (\@FAO20.

@@

-   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests ( @poorter2016 ), and offset \~10% of carbon emissions from Amazonian deforestation ( @smith2020 ).
-   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y ( @rozendaal2019 ).
-   However, secondary neotropical forests diverge widely in taxonomic composition ( @norden2015 ).

@@

Half of all forests lie half of a kilometer from an ecotone ( @haddad2015 ), suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.

-   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass ( @ordway2020 ).

@@

@@

Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).

@@

Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).

@@

Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),

with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.

@@

with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),

>>>>

---
title: "Introduction"
output: html_notebook
bibliography: ../../library.bib
---

# Outline

1.  importance of reforestation/regeneration/regrowth
2.  patterns of reforestation
    1.  general
    2.  tropical
3.  things that affect reforestation
    1.  edge effects
    2.  plantation structure?
4.  [for NGO--start biomass census]
    1.  Does leaving the plantation structure help reforestation?
        1.  shade
        2.  edge effects (i.e. buffer)
    2.  How does the forest regenerate by distance to primary forest?
        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?

# Draft

@@

    - Tropical forests store most terrestrial biomass *[@pan13]* and biodiversity (CITE, but are losing functions, especially in the neotropics *[@hubau2020]*, as most tropical forests become secondary *[@FAO20]*.

@@

    - Neotropical secondary forests accumulate biomass >10x faster than old-growth forests ( @poorter2016 ), and offset ~10% of carbon emissions from Amazonian deforestation *[@smith20]*.
    - They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y *[@rozendaal19]*.
    - However, secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*.

@@

Half of all forests lie half of a kilometer from an ecotone *[@haddad15]*, suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.

    - Tropical forests lose 22% carbon along edges >100m, in part due to shorter trees and less leaf mass *[@ordway20]*.

@@

# attic

@@

Syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years), mainly with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions *[@rozendaal19; @onyekwelu16]*.

@@

Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests *(@nordon15, Christensen 2014, Munroe et al 2013, Chazdon et al 2007)*.

@@

Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management *[@wright09]*,

with underlying arguments based on metapopulation theory in ecology *[@levins69; @macarthur67]*, but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.

@@

with examples of high diversity (compared to older adjacent stands) in agroforests *[@oliveira-neto17]* and native timber plantations (Pryde et al 2015),

### 2022-04-25 article/methods.Rmd
draft res,fig3 bug

---
title: "Methods"
output: html_notebook
---

@@

**See () for an ecosystem description.**

@@

This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig 1)* and a wide service road on the third NE border, with primary forest beyond it.

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig 1)*, using GIS software *(ArcGIS, QGIS)*.

@@

The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *(Onyekwelu & Olabiwannu 2016)*.

## Plot measurements / Data collection

### Canopy cover & Topography

@@

Tree height was recorded by measuring distances to both the crown and the stem at chest height (~2.7 m) using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and trangulating the missing side length.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by *Chave et al. (2014)*.

The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by *Chave et al. (2005) (Ngomanda et al 2014, Alvarez et al 2012)*.

@@

The equation used was:
\
*AGB = 0.0673 x (ρ D^2^ H)^0.976^*
\
with *D* in cm, *H* in m, and *p* in g cm^-3^.

@@

Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* ( `r `packageVersion("BIOMASS")` ) R software package by *Réjou-Méchain et al. (2017)*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

@@

For all response variables, data were analyzed using plot medians and subject to linear regression with the distance to forest edge variable after checks for residual normality and equal variance assumptions using R `getRversion()` *(R Core Team 2019)*.

@@

Non-linear regressions were run using the function *poly()* in the *stats* `packageVersion("stats")` R package *()*, 
which was only recorded for stem density and richness variables.

@@

Community analysis and ordination was done as a PERMANOVA with the *vegan* `packageVersion("vegan")` R package *()*.

@@

>>>>

---
title: "Methods"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: false
bibliography: ../../library.bib
---

```{r echo=F, results='hide', warning=F}
knitr::opts_chunk$set(echo = F, warning=F)
library(here)
```

@@

(See *@taylor15* for a broader ecosystem description of the region.)

@@

This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides *(Fig \@ref(fig:map))* and a wide service road on the third NE border, with primary forest beyond it.

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E *(Fig \@ref(fig:map))*, using GIS software *(ArcGIS, QGIS)*.

@@

The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

```{r map, fig.cap="Map of conservation area on Osa Peninsula, Costa Rica."}
knitr::include_graphics(here("figs/map.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```
<!-- ![**Fig 1**](../figs/map.png) -->

## Plot measurements 
<!-- / Data collection -->

### Canopy cover and topography

@@

Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.

@@

Aboveground biomass (AGB) was estimated per stem using the allometric equation by ... *[@chave15]*.

The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by ... *[@chave05] (Ngomanda et al 2014, Alvarez et al 2012)*.

@@

The equation used was:

*AGB = 0.0673 x (ρ D^2 H)^0.976*

with *D* in cm, *H* in m, and *p* in g cm^-3.

@@

Original values from 2013 were later updated to modern values using the pipeline developed in the *BIOMASS* `r packageVersion("BIOMASS")` R software package by ... *[@réjou-méchain17]*, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

@@

For all response variables, data were analyzed using plot medians and subject to linear regression with the distance to forest edge variable after checks for residual normality and equal variance assumptions using `r version$version.string`.

@@

Non-linear regressions were run using the function *poly()* in the *stats* `r packageVersion("stats")` R package *(R Core Team 2022)*, 
which was only recorded for stem density and richness variables.

@@

Community analysis and ordination was done as a PERMANOVA with the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.

@@

<!-- `r citation()` -->

### 2022-04-25 article/results.Rmd
draft res,fig3 bug

---
title: "Results"
output: html_notebook
---

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(
  echo = F,
  message = F,
  warning = F
)
library(here)
```

## Biomass

Distance to primary forest edge was associated with ### more ___, although this did not correlate with ___.

@@

```{r fig1}
source(
  here(
    "figs/fig1.R"
  )
)
fig1
```

## Composition

@@

```{r fig2}
source(
  here(
    "figs/fig2.R"
  )
)
fig1
```

>>>>

---
title: "Results"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: false
---

```{r setup, echo = F, message = F, warning=F, results=F}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(here); i_am("article/results.Rmd"); source(here("figs/fig2.R"))
```

# Biomass

@@

```{r fig1data}
wd <- filter(plotResultsTbl1, variable == "spg") #alias
stems <- filter(plotResultsTbl1, variable == "stems") #alias
stems12 <- filter(plotResultsTbl12, variable == "stems") %>% distinct() #alias
```

Aboveground biomass did not tend to change with distance to primary forest edge 
*(Fig \@ref(fig:biomass)a)*, despite significant trends with underlying related variables 
*(Fig \@ref(fig:biomass))*.

Median plot wood density tended to increase marginally significantly 
(`r glue("p=",round(as.numeric(wd$pval),3))`) 
with edge distance by 
`r round(wd$statPrint[[1]]$estimate[[2]],4)` ± 
`r round(wd$statPrint[[1]]$std.error[[2]],5)` per m 
*(Fig \@ref(fig:biomass)b)*.

Plot stem density also tended to increase significantly 
(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
with edge distance but more strongly and non-linearly, by 
~`r round(stems12$statPrint[[1]]$estimate[[2]],2)` ± 
`r round(stems12$statPrint[[1]]$std.error[[2]],2)` per m 
*(Fig \@ref(fig:biomass)c)*.

At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
*(Fig \@ref(fig:biomass)d)*.

```{r biomass, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability."}
knitr::include_graphics("../figs/fig1.png")
```

# Composition

@@

```{r fig2data}
entropy <- filter(plotResultsTbl1, variable == "entropy") #pReplaced
entropy12 <- filter(plotResultsTbl12, variable == "entropy") %>% distinct() #slopeR2
richness <- filter(plotResultsTbl1, variable == "richness")
ficus <- filter(taxaResultsTbl1, gen == "Ficus")
```

Overall diversity changed significantly with distance to primary forest edge 
*(Fig \@ref(fig:taxa))*.

Alpha diversity significantly 
(`r glue("p=",round(as.numeric(entropy12$pval[[2]][2]),3))`) 
decreased non-linearly by 
~`r round(abs(entropy12$statPrint[[1]]$estimate[[2]]),2)` ± 
`r round(abs(entropy12$statPrint[[1]]$std.error[[2]]),2)` per m 
as distance to primary forest edge increased 
which explained 
`r "R2x100"`% of variance among distance strata medians 
*(Fig \@ref(fig:taxa)a)* -- 
while in contrast taxonomic richness increased slightly 
(`r glue("p=",round(as.numeric(richness$pval),3))`) 
and linearly 
by 
`r round(richness$statPrint[[1]]$estimate[[2]],2)` ± 
`r round(richness$statPrint[[1]]$std.error[[2]],3)` per m 
and had `r "R2x100"`% variance explained among distance strata 
*(Fig \@ref(fig:taxa)b)*.

Community composition and beta diversity also changed significantly 
(`r glue("p=",round(ordStat$aov.tab[[6]][1],2))`) with 
`r round(ordStat$aov.tab$R2[1],2) *100`% 
variance among distance strata explained by distance to primary forest edge, 
and the first two principal components explaining 
`r ordTbl$PC1_prop[1] *100`% and 
`r ordTbl$PC2_prop[1] *100`% totaling 
`r (ordTbl$PC1_prop[1] + ordTbl$PC2_prop[1]) *100`% 
of variance among calculated plot distances 
*(Fig \@ref(fig:taxa)d)*.

Key abundant taxa `r measure2e` and `r measure2f` showed different responses -- `r measure2e` nearly tended to decrease with distance to primary forest edge 
*(Fig \@ref(fig:taxa)e)*,
while `r measure2f` decreased marginally significantly 
(`r glue("p=",round(as.numeric(ficus$pval),3))`) 
and linearly by 
`r round(ficus$statPrint[[1]]$estimate[[2]],2)` ± 
`r round(ficus$statPrint[[1]]$std.error[[2]],2)` per m 
with `r "R2x100"`% variance among distance strata explained 
*(Fig \@ref(fig:taxa)f)*.

```{r taxa, fig.cap=" Taxonomic composition measures including (a) diversity, (b) richness, (c) overall biomass, (d) beta diversity, and (e,f) biomass of key taxa."}
knitr::include_graphics("../figs/fig2.png")
```

# Functions

```{r fig3data}
source(here("figs/fig3.R")); succession <- taxaResultsTbl12 #pReplaced
```

Overall functional regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
*(Fig \@ref(fig:traits))*.

Taxa associated with both early and late successional stages decreased significantly 
(`r glue("p=",round(as.numeric(succession$pval[[3]]),3))`) 
and curvi-linearly 
by 
~`r round(succession$statPrint[[1]]$estimate[[2]],2)` ± 
`r round(succession$statPrint[[1]]$std.error[[2]],2)` per m 
with distance to edge explaining 
`r "R2x100"`% variance among distance strata medians 
*(Fig \@ref(fig:traits)a)*.

Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
*(Fig \@ref(fig:traits)b)*.

```{r traits, fig.cap=" Functional regeneration based on taxa associations with (a) successional stages and (b) primary dispersal mode."}
knitr::include_graphics("../figs/fig3.png")
```

### 2022-04-25 talk/talk.Rmd
draft res,fig3 bug

```{r setup, echo = F, include = F}
knitr::opts_chunk$set(
  echo = F,
  quiet = T
)
library(
  here
)
i_am(
  "talk/talk.Rmd"
)
# #fixRender
# library(
#   httpuv
# )
#embed_file()
# library(
#   xfun
# )
# xfun::pkg_load2(c("htmltools", "mime"))
```

@@

>>>>

```{r setup, echo = F, include = F}
knitr::opts_chunk$set(echo = F, quiet = T)
library(here); i_am("talk/talk.Rmd")
# #fixRender
# library(
#   httpuv
# )
#embed_file()
# library(
#   xfun
# )
# xfun::pkg_load2(c("htmltools", "mime"))
```

@@

- Edge effects may shape diversity via dispersal overlap, possibly peaking at intermediate values *[@yamaguchi22]*
